---
title: "MFE R Programming Workshop"
author: "Brett Dunn and Mahyar Kargar"
date: "Fall 2017"
output:
  beamer_presentation:
    includes:
      in_header: ../slide_header.tex
    keep_tex: false
subtitle: Week 4
slide_level: 2
---
# Introduction

## Questions
\begin{center}
\Large Any questions before we start?
\end{center}

## Overview of Week 4
- Strings
- Dates
- Lab

# Strings

## Strings
- A string is a sequence of characters.
- In `R`, a sting falls in the `character` class.
```{r}
mystring <- "Hello"
str(mystring)
```
- Character vectors are created like numeric vectors.
```{r}
myvec <- c("Hello", "Goodbye")
str(myvec)
```

## Manipulating Strings
- `R` provides many functions to manipulate strings.
    - `grep()`: Searches for a substring, like the Linux command of the same name.
    - `nchar()`: Finds the length of a string.
    - `paste()` and `paste0()`: Assembles a string from parts.  
    - `sprintf()`: Assembles a string from parts.
    - `substr()`: Extracts a substring.
    - `strsplit()`: Splits a string into substrings.
- Hadley Wickham's `stringr` package provides additional functions for using regular expressions and examining text data.

## `grep()`
- The call `grep(pattern,x)` searches for a specified substring pattern in a vector `x` of strings.
- If `x` has `n` elements—that is, it contains `n` strings—then `grep(pattern,x)` will return a vector of length up to `n`.
-Each element of this vector will be the index in `x` at which a match of pattern as a substring of `x` was found.
```{r}
grep("Pole",c("Equator","North Pole","South Pole"))
```

## `nchar()`
- The call `nchar(x)` finds the length of a string `x`.
```{r}
nchar("South Pole")
```

## `paste()`
- The call `paste(...)` concatenates several strings, returning the result in one
long string.
```{r}
paste("North","and","South","Poles")
paste("North","Pole",sep="")
# paste0 is same as sep="" (more efficient)
paste0("North","Pole") == paste("North","Pole",sep="") 
```

## `sprintf()`
- The call `sprintf(...)` assembles a string from parts in a formatted manner.
- Similar to the `C` function `printf`.
```{r}
i <- 8
sprintf("the square of %d is %d",i,i^2)
```

## `substr()`
- The call `substr(x,start,stop)` returns the substring in the given character position range `start:stop` in the given string `x`.
```{r}
substring("Equator",3,5)
```

## `strsplit()`
- The call `strsplit(x,split)` splits a string `x` into an `R` list of substrings based on another string `split` in `x`.
```{r}
strsplit("10-05-2017",split="-")
```

## Example: Creating File Names
- Suppose we want to create five files, `q1.pdf` through `q5.pdf`, consisting of
histograms of 100 random N(0,i2) variates. We could execute the following
code:
```{r, eval=FALSE}
for (i in 1:5) {
  fname <- paste("q",i,".pdf")
  pdf(fname)
  hist(rnorm(100,sd=i))
  dev.off()
}
```

# Dates


## Why do we need date/time classes?

\centerline{
  \includegraphics[width=\textwidth,height=0.85\textheight,keepaspectratio]{./railroad_time.jpg}
}


## Date Classes in R
- `Date` is in `yyyy-mm-dd` format and represents the number of days since Jamuary 1, 1970
- `POSIXct` represents the (signed) number of seconds since Jamuary 1, 1970 (in the UTC time zone) as a numeric vector.
- `POSIXlt` is a named list of vectors representing `sec`, `min`, `hour`, `mday`, `mon`, `year`, time zone par maters, and a few other items.
```{r}
x <- Sys.time()  # clock time as a POSIXct object
x; as.numeric(x)
```

## Creating Dates
- Typically, dates come into R as character strings.
- By default, R assumes the string is in the format `yyyy-mm-dd` or `yyyy-mm-dd`
```{r}
mychar <- "2017-10-05"
mydate <- as.Date(mychar)
str(mydate)
```

## Date Formats
- `R` can parse many other types of date formats.
- See `?strptime` for details.
```{r}
mychar <- "October 5th, 2017"
mydate <- as.Date(mychar, format = "%B %eth, %Y")
str(mydate)
```

## Extract Parts of a Date Object
```{r}
mydate <- as.Date("2017-10-05")
weekdays(mydate)
months(mydate)
quarters(mydate)
```

## Generate Regular Sequences of Dates
```{r, tidy=TRUE, size = "footnotesize"}
## first days of years
seq(as.Date("2007/1/1"), as.Date("2010/1/1"), "years")
## by month
seq(as.Date("2000/1/1"), by = "month", length.out = 4)
## quarters
seq(as.Date("2000/1/1"), as.Date("2001/1/1"), 
    by = "quarter")
```

## Time Intervals / Differences
- Function difftime calculates a difference of two date/time objects and returns an object of class "difftime" with an attribute indicating the units.
```{r}
time1 <- as.Date("2017-10-05")
time2 <- as.Date("2008-07-08")
time1 - time2
difftime(time1, time2, units = "weeks")
```


## Dates in Microsoft Excel
- Microsoft Excel stores dates as the number of days since Decemeber 31, 1899.
- However, Excel also incorrectly assumes that the year 1900 is a leap year to allow for compatability with Lotus 1-2-3.
- Therefore, for dates after 1901, set the origin to Decemeber 30, 1899 to convert an Excel date to an R date.
```{r}
as.Date(43013, origin = "1899-12-30")
```

# Lubridate

## Lubridate
 
- Lubridate is an R package that makes it easier to work with dates and times.
- Lubridate was created by Garrett Grolemund and Hadley Wickham.

```{r}
# install.packages("lubridate")
library(lubridate)
```
 

## Parse a date
 
- Lubridate accepts lots of formats

```{r}
ymd("20110604")
mdy("06-04-2011")
dmy("04/06/2011")
```
 

## Parse a date and time

```{r}
ymd_hms("2011-06-04 12:00:00", tz = "Pacific/Auckland")
```
 

## Extraction

```{r}
arrive <- ymd_hms("2011-06-04 12:00:00")
second(arrive)
second(arrive) <- 25
arrive
```
 

## Intervals

```{r}
arrive <- ymd_hms("2011-06-04 12:00:00")
leave <- ymd_hms("2011-08-10 14:00:00")
interval(arrive, leave)
```
 

## Arithmetic

```{r}
mydate <- ymd("20130130")
mydate + days(2)
mydate + months(5)
```
 

## Arithmetic
```{r}
mydate <- ymd("20130130")
mydate + days(1:5)
```
 

## End of (next) month

```{r}
jan31 <- ymd("2013-01-31")
jan31 + months(1)
ceiling_date(jan31, "month") - days(1)
floor_date(jan31, "month") + months(2) - days(1)
```

## Lab 2
\begin{center}
\Large Let's work on Lab 2.
\end{center}